Periodic Geodesics on the Infinite Necker Cube Surface

### 2018-01-04 newpaper.tex
010318

\title{Periodicity of Geodesics on the Necker Cube Surface}
\author{Pavel Javornik}

@@

\begin{abstract}
\noindent Although it is most recognizable as M.C. Escher's adaptation of an optical illusion popularized by Louis Albert Necker, the Necker cube surface has a complex structure and is, in analaytic language, described as an infinite, flat (locally isometric to $\RR^2$ on smooth sections), $\RR^3$-embeddable Euclidean cone surface with countably many singularities of cone angles $3\pi$ and $\frac{3\pi}{2}$. Its geometric construction is described as the gluing of infinitely many unit cubes along their edges so that every face shares an edge with exactly four others. A unit-speed geodesic on this surface has an initial trajectory angle $\theta\in\RR/2\pi\ZZ$ given up by the surface's local plane isometry. Given a rational vector direction in $\RR^2$ (components are relatively prime integers) obtained from $\theta$, one can classify every periodic/drift-periodic rational geodesic on the surface. In addition its total period/length and, in the case where it is drift-periodic, translational drift as Euclidean distance in its ambient space is given by solving a system of simple linear Diophantine equations. 
\end{abstract}
\begin{figure}[H]
\centering
\includesvg[width=4in]{cubesxyzpaths2bwv2b}
\label{fig:front}
\caption{Periodic and drift-periodic flows on the Necker cube surface}
\end{figure}

\newpage
\section{Introduction}

Put your Problem statement here! Example of a Citation\cite[p.219]{Robotics}. Here's Another Citation \cite{Flueck}

The Necker cube has made numerous appearances in the work of mathematicians, crystallographers, and scientists interested in human visual systems before and after appearing in the works of M.C. Escher [cite]. Louis Albert Necker was credited for having discovered the optical illusion, and studying its geometric properties [cite]. The solid presentation of the cube, when rendered as a flat surface with three rear faces removed, achieves a similar effect when its three visible faces are shaded in a particular way. The Necker cube surface is then obtained by gluing infinitely many copies along its outer edges as in Figure 1.

Dynamical classification of geodesics on flat, periodic surfaces such as the Necker cube surface ultimately boils down to  studying the surface's various symmetries and how the rotational holonomy group acts on the unit tangent bundle of the surface as the unit vector (derivative of the geodesic) is parallel transported along its path. Loops on polyhedra punctured at its cone singularities tend to have discrete holonomy groups as its relative holonomy (holonomy of contractible loops) is trivial.

\begin{rem}
Familiarity is assumed on the part of the reader with covering space theory, translation (Veech) surfaces, and their associated Veech groups. For general surveys on these topics: [cite], [cite].
\end{rem}

\subsection{Discussion of Results}
Our initial experiments strongly supported the theory that there would be a correlation between a choice of trajectory angle and dynamical properties of a geodesic on such a symmetric object. Rightfully so, a surface composed of infinitely many cubes an induced flat metric where every neighborhood is locally isometric to the Euclidean plane. Via the parallel transport of a unit tangent vector over the sharp edges of the surface denoted $\bUs$, it becomes necessary to think of the geodesic as a sequence of line segments contained in the faces of each cube on the surface's embedded form in $\RR^3$. $\ref{fig:front}$.
\\
\begin{wrapfigure}{l}{2in.}
\includesvg{vectorplane}
\end{wrapfigure}
\noindent Let $u_0$ be a point on the surface that is contained in the interior of a face and consider a tangent unit vector $v\in\RR^3$ based at point $u_0$. Each face is parallel to a $2$-dimensional subspace of $\RR^3$ spanned by exactly two basis vectors, so one component of $v$ is 0. Call this projected vector $v_0\in\RR^2$.
\\\\\\
The angle $v_0$ makes relative to a choice of basis is the \emph{initial trajectory angle} of the unit-speed geodesic on the surface, $\Phi_t:\RR\rightarrow\bUs$. A rational initial trajectory angle in $\RR^2$ falls into one of two categories:
\begin{Def}
Let $v_0$ be a unit vector of the form $\frac{1}{k}(x,y)\in\RR^2$ with $x,y\in\mathbb{Z}$ and $k=\sqrt{x^2+y^2}\in\mathbb{R}$. We say $v_0$ is an \textbf{odd-odd} vector if its components are relatively prime and both odd. We denote the \textbf{set of all odd-odd directions} by $\mathcal{O}$. We say that $v_0$ is an \textbf{even-odd} vector if its components are relatively prime and of opposite parity. We denote the \textbf{set of all even-odd directions} by $\mathcal{E}$.
\end{Def}
\noindent For convenience we call the disjoint union of these sets $\cS$. This paper aims to depict the relationship between initial trajectory angles and dynamical properties of a geodesic on the surface. Denote $\bUs$'s translational isometries by $\text{Trans}(\bUs)$. Recall that a geodesic on the surface is \emph{periodic} if there exists some $T>0$ such that $\Phi(t+T)=\Phi(t)$ for all $t\in\RR$ and a geodesic is \emph{drift-periodic} if there exists some $T'>0$ and non-trivial $p\in \text{Trans}(\bUs)$ such that $\Phi(t+T')=p(\Phi(t))$ for all $t\in\RR$. In either case we call this value the \emph{period} of $\Phi$ if it is the \emph{smallest} possible value that satisfies our definition. We additionally require that at no point does $\Phi$ encounter a cone singularity of the surface, in which case the geodesic behavior is non-deterministic.

\begin{align}
\text{Let }\XX'=\left\< \left[ \hspace{1mm} \begin{matrix}
							1 & 2\\
							 0 & 1
						\end{matrix}\hspace{1mm}\right],~
\left[ \hspace{1mm} \begin{matrix}
							1 & 0\\
							 2 & 1
						\end{matrix}\hspace{1mm}\right]\right\>
\end{align}

\noindent Section [blank] explains the significance of this group. Let $M=\left[ \hspace{1mm} \begin{matrix}
							s & u\\
							 w & z
						\end{matrix}\hspace{1mm}\right]\in\XX'$.

\begin{thm*}
(Dynamical Properties of a Geodesic on $\bUs$) \\Let $\Phi_t:\RR\rightarrow\bUs$ be a non-singular unit-speed geodesic on the Necker cube surface such that $\Phi(0)=u_0$ and $\Phi'(0)=v\in\RR^3$, isometric to $v_0\in\RR^2$. Let $\overline{u}_t\in\RR^3$ be $\Phi_t$'s coordinate in $\bUs$'s ambient space at time $t$. Then the following is true:
\begin{enumerate}[label=(\roman*)]
\item $\Phi$ is periodic if $v_0\in\cO$.
\item Suppose $(i)$ is true. If $M (kv_0)=(1,1)$ then $\Phi$ has period $T=6\sqrt{2}||kv_0||$.
\item $\Phi$ is drift-periodic if $v_0\in\cE$. 
\item Suppose $(iii)$ is true. If $M (kv_0)=(1,0)$ or $M (kv_0)=(0,1)$ then $\Phi$ has period $T=2||kv_0||$, and $d_E(\overline{u}_t,\overline{u}_{t+T})=||kv_0||\sqrt{2}$.
\end{enumerate}
\end{thm*}
\compav{Remember to prove that such an M exists.}

Here $||~||$ is the standard Euclidean norm in $\RR^2$, and $d_E:\RR^3\times\RR^3\rightarrow\RR$ is the Euclidean distance. The proof for this theorem can be found in XXX and follows from Theorems YYYY and ZZZZZ.

@@

\section{Periodic Tiling of Necker Cubes}
This section will detail how the Necker Cube surface is constructed as a covering space of a thrice-punctured torus. Make the following identifications on 3 unit squares:
\begin{figure}[H]
\centering
\includesvg[width=1.83in]{neckercube}\includesvg[width=2in.]{neckercubepath}
\caption{$\bGs$ with and without paths $a,b,c,d$.}
\end{figure}

We will call this ``half-cube" $\bG$. It is a genus one piecewise smooth surface homeomorphic to $\TT^2$. Every vertex has a cone angle of either $3\pi$ or $\frac{3\pi}{2}$. Let $\Sigma_{\bG}\subset\bG$ be the set of these three vertices. These are conical singularities of the surface. We denote the surface punctured at these points by $\bG\backslash\Sigma_{\bG} = \bGs$. The paths labeled $a,b,c,d$ are independent of one another and span $\bGs$'s first fundamental group, $\pi_1(\bGs)\cong \<a,b,c,d \>$, the free group of four generators. The unpunctured Necker cube surface $\bU$ is embedded in $\RR^3$ by tiling along $\bG$'s outer edges:

\begin{figure}[H]
\centering
\includesvg[width=4in]{neckercubepathsurface}
\end{figure}
From this diagram we see that $\bU$ is a universal cover of $\TT^2$. In this paper we are studying geodesic behavior on $\bU$ and in order to do so, it is necessary to determine how the parallel transport of a unit vector around a cone singularity acts on the unit tangent bundle of the surface. But to study the path monodromy we require that these loops (see $c,d$ in Fig 3) aren't contractible on the cube's sharp corners. Denote the countably infinite set of singularities of $\bU$ by $\Sigma_{\bU}$. The \emph{Necker cube surface} is the punctured surface $\bUs=\bU\backslash\Sigma_{\bU}$. Observe from the previous diagram that $\pi_1(\bUs)$ is the kernel of the following group homomorphism:

\begin{Def}
Denote $\varphi_1:\pi_1(\bGs)\rightarrow \ZZ^2$ as the group homomorphism such that $c,d\mapsto(0,0)$, $a\mapsto(1,0)$, and $b\mapsto(0,1)$.
\end{Def}

Take $\pi_1(\bUs)$ to be $\ker\varphi_1$. Then $\Delta_{\varphi_1}=\pi_1(\bGs)/\pi_1(\bUs)\cong\ZZ^2$ is the \emph{deck group} of the covering map $\varphi_1^*:\bUs\rightarrow\bGs$. Now $\bGs$ is unfolded and flattened in order to study monodromy as it transports a vector in $\RR^2$ along a path. Apply the following cutting and gluing operations to the surface: \\
\begin{figure}[H]
\centering
\includesvg[width=3in]{LTorus}
\end{figure}
\noindent This L-shaped torus is then taken apart and recovered as $2\times 2$ torus missing a unit square with the following identifications:\\
\begin{figure}[H]
\centering
\includesvg[width=2in]{BranchTorus}
\caption{$\bGs$ and images of paths after reconstruction. The outermost vertex is \emph{not} a puncture.}
\end{figure}

Consider a $2\times2$ sheet missing an open unit square and that square's vertices (punctures). Center this structure around $0\in\CC$ and recover $\bGs$ with the identifications made in Fig 3. Let $m,n\in\ZZ$. Similar identifications can be made on the following subset of $\CC$ to recover $\bUs$:

@@

\noindent Let $\aleph_{m,n}=2m+i2n\in\CC$ be the center of a removed square. Now let $\zeta_{m,n}(t)=(2m+t)+i(2n-\frac{1}{2})$ and 
$\xi_{m,n}(t)=(2m+t)+i(2n+\frac{1}{2})$ be paths on $\mathbf{P}$ that run along the edges of the squares, parameterized by $t\in (-\frac{1}{2},\frac{1}{2})$ and $(m,n)\in\ZZ^2$. $\bUs$ is recovered as a topological quotient $\mathbf{P}/\sim$, where $\sim$ is the minimal relation on $\mathbf{P}$:

\begin{equation}
\begin{split}
\zeta(t)\sim e^{i\frac{\pi}{2}}~\overline{\zeta(t)-\aleph}\\
\eta(t)\sim e^{i\frac{\pi}{2}}~\overline{\eta(t)-\aleph}
\label{eq:rel2}
\end{split}
\end{equation}
The translations, $\pm 2,\pm 2i$, on $\mathbf{P}$ are isometries that preserve these identifications and induce isometries of $\bUs$. The group generated by these horizontal and vertical translations form the \emph{group of Deck transformations}, $\text{Deck}(\varphi_1^*)$, that  is isomorphic to $\ZZ^2$:

@@

Identify 2 with $f_h$, and $2i$ with $f_v$. Let $x_0\in\bGs$ and fix a point in the fiber,  $u_0\in\varphi_1^*{-1}(x_0)\subset\bUs$. The action of $\ZZ^2$ on the fiber over $x_0$ is given by $(m,n)\cdot u_0=(f_h^m\circ f_v^n)(u_0)$. The distance between $u_0$ and $(f_h^m\circ f_v^n)(u_0)$ is measured in its ambient space and taken to be the magnitude of $2m+i2n\in\CC$.

\begin{lem}
Let $[\alpha]$ be homotopy class not in the kernel of $\varphi_1$. It will lift to an unclosed path terminating on $\varphi_1([\alpha])\cdot u_0$.
\end{lem}

\subsection{Monodromy Group Representation}
A \emph{rotational monodromy map} is a morphism from the first fundamental group of a surface to the \emph{monodromy group} acting on the unit tangent bundle via parallel transport of a vector along its path. Existence of a parallel transport map is a consequence of the metric connection on a surface with complex structure. The monodromy map is well defined when the \emph{holonomy group} is \emph{discrete}. When the \emph{relative holonomy} of contractible loops is trivial, the holonomy group is the set of attainable rotations of a unit vector upon returning to the base point after transporting it along a non-trivial loop. A \emph{flat surface} like $\bGs$ has trivial relative holonomy as its neighborhoods are isometric to the plane. Denote the \emph{unit tangent bundles} of $\bUs$ and $\bGs$ by $T^1\bUs$ and $T^1\bGs$. If $\alpha$ is a loop on $\bGs$, then $[\alpha]$ acts on a fixed vector $v_0\in\RR^2$ by rotation. In the case where $\alpha$ is \emph{not} a geodesic segment on the surface of length $L$, $v_0=\alpha'(0)\in\RR^2$ may differ from $\alpha'(L)$ by some rotation.

\begin{Def}
Define $\varphi_2:\pi_1(\bGs)\rightarrow SO(2,\ZZ)$ to be the group homomorphism where $a,b \mapsto I_2$, $c  \mapsto R$, and $d  \mapsto R^3$, where $I_2$ is the identity matrix and $R$ is a positive $90^\circ$ rotation matrix.
\end{Def}

\begin{figure}[H]
\centering
\includegraphics[width=2.5in]{monodromy.png}
\caption{Effect a nontrivial loop has on basis vectors $(i,j)\in\RR^2$.}
\label{fig:loop}
\end{figure}

\begin{lem}
$\varphi_2:\pi_1(\bGs)\rightarrow \mathrm{SO}(2,\ZZ)$ is the rotational monodromy representation of $\bGs$.
\begin{proof}
From Figure 4 it's clear that $c,d$ will rotate a unit vector by $\pm 90^\circ$ by parallel transport along their paths. These values are determined by the cone angles of the singularities they loop around. The parallel transport of a vector along $a$ or $b$ would be trivial since the outer edges they pass through are glued by translation.
\end{proof}
\end{lem}

\begin{lem}
Any element in $\text{Deck}(\varphi_1^*)$ acting on $T^1\bUs$ has a trivial effect on holonomy.
\begin{proof}
Since $\bGs=\bUs/\ZZ^2$, the deck group action lifts naturally to an action on their tangent vector spaces and unit tangent bundles via pushforward of deck transformations $f_h,f_v$ on the bundle maps. i.e. $T^1\bGs=T^1\bUs/\ZZ^2$. This effect must be trivial since $\varphi_1(a)\sim f_h$ and $\varphi_1(b)\sim f_v$, and both homotopy classes $a,b$ have trivial holonomy.  \\
How do $f_h,f_v$ parallel transport a vector?\\
Well if $\varphi_1([a])\cdot u_0=f_h(u_0)$, then for 
\end{proof}
\end{lem}

\begin{thm}
Let $m=\varphi_2\vert_{\pi_1(\bUs)}$. Then $m$ is the rotational monodromy map of $\bUs$.
\begin{proof}

\begin{cor}
Let $\alpha:[0,1]\rightarrow\bGs$ be a loop on $\bGs$ with trivial holonomy and $\tilde{\alpha}$ its lift to $\bUs$. Then $\tilde{\alpha}$ has trivial holonomy as well.
\begin{proof}

\end{proof}
\end{cor}

>>>>

\setlength{\topmargin}{1in} %top/bot margins
\setlength{\oddsidemargin}{\topmargin} %sidemargins

\setlength{\textheight}{11in} \setlength{\textwidth}{8.5in}
\setlength{\hoffset}{-1in} \setlength{\voffset}{-1in} \setlength{\evensidemargin}{\oddsidemargin} \addtolength{\textheight}{-2 \topmargin}\addtolength{\textwidth}{-2\oddsidemargin}
\setlength{\headheight}{0pt} \setlength{\headsep}{20pt} \setlength{\footskip}{20pt}
\addtolength{\textheight}{-\footskip} \addtolength{\textheight}{-\headheight} \addtolength{\textheight}{-\headsep}

\title{Periodic Geodesics on the Infinite Necker Cube Surface}
\author{Pavel Javornik}

@@

\begin{abstract}
\noindent We classify all of the periodic and drift-periodic trajectories on the surface periodically constructed out of unit cubes lying on a plane in $\RR^3$. The Necker cube surface is a piecewise smooth Euclidean cone surface with neighborhoods locally isometric to the plane that gives up a natural piecewise isometry, or \emph{flattening}, onto a subset of $\CC$ with $\ZZ^2$ translational symmetries. We use the surface's topological properties to draw a connection between translation covers and rotational holonomy of closed curves by inducing fundamental group actions on the discrete space of attainable vectors via parallel transport along lifted paths. We then use the symmetries of this surface and its (branched) translation cover to show a correspondence between the space of attainable (rational) geodesics of the Necker cube surface, their dynamical properties, and the translation cover's \emph{Veech group}. The ``slope" of a geodesic is represented by a relatively prime pair of integers, which is used to tell not only whether or not a geodesic with this initial direction (relative to choice of basis in its initial tangent space) closes, but the overall length of the trajectory as a function of its components.
\end{abstract}
\begin{figure}[H]
\centering
\includesvg[width=4in]{cubesxyzpaths2bwv2b}
\label{fig:front}
\caption{Periodic and drift-periodic geodesics on the Necker cube surface}
\end{figure}
\newpage
\section{Introduction}
The Necker cube has been studied extensively due to its interesting effects on human visual perception. \cite{Albert} A popular adaptation of the Necker cube is the edge-to-edge pairing of multiple solid cubes giving the illusion of a rhombile tiling of the plane when viewed from a fixed angle. (Fig 1) It is most recognizable in the works of M.C. Escher \cite{Escher1}\cite{Escher2} and as the game board in the 80s cabinet arcade classic, $\text{Q}^*\text{bert}$.\cite{Qbert} The cube was named after the Swiss crystallographer Louis A. Necker for having been one of the first people to study the wire-frame cube's properties as an optical illusion.

 The solid presentation of the cube, rendered as a flat surface with three rear faces obscured from view, achieves a similar effect to the ``wire-frame" version when infinitely many are arranged along their outer edges. The semi-cube is a construction obtained by removing these obscured faces and making translational identifications on these edges that is homeomorphic to the two-dimensional torus. The Necker cube surface is itself obtained by gluing infinitely many copies of the semi-cube surface along those outer edges. (Fig ??)

An infinite surface like the Necker cube is piecewise smooth and \emph{flat} on its smooth sections. Neighborhoods contained in these faces are isometric to the Euclidean plane. The tangent space at a point is the vector field contained in a plane tangent to the surface. A vector field translated along a closed loop on this surface induces a phenomena known as \emph{rotational holonomy}. This is a linear transformation (element of $\SO(2,\RR)$ for orientable manifolds) between tangent spaces at the initial and terminal point of the loop. When a surface is flat, deformations of the loop do not alter its holonomy. Furthermore, a contractible (trivial) loop has trivial holonomy due to the local plane isometry on its smooth sections. In this case any homotopic path will induce the same linear transformations, and that gives up a well-defined group action on the tangent bundle of the surface by the first fundamental group of the surface.

In studying geodesics on an infinite flat surface, it is important to make this connection early on. The rotational holonomy group acts transitively on the holonomy bundle, a principal bundle of tangent spaces over a point on the surface. This holonomy group is used to construct a covering space where the only closed loops are those with trivial holonomy. In desirable cases, the holonomy group is discrete and this cover has finitely many sheets. This surface is known as an infinite \emph{translation surface}. It is on this surface that geodesics behave as they would when projected back down to the flat surface. In the context of Riemann surfaces, a translation surface is a surface with a maximal translational atlas, i.e. every transition function on images of neighborhoods in $\CC$ is an affine translation. The surface inherits a flat metric from pulling back the Euclidean metric on $\CC$. On such a surface the principal holonomy bundle of tangent spaces over a point consists of a single vector space. Combining these two facts, in local coordinates any geodesic in direction $\theta\in\RR/2\pi\ZZ$ on this surface is isometric to $te^{i\theta}$ for $t\in\RR$. [cite Zorich]
\begin{rem}
Familiarity is assumed on the part of the reader with covering space theory, translation (particularly Veech) surfaces, and their associated Veech groups. For general surveys on these topics we refer readers to: [cite], [cite].
\end{rem}

\subsection{Directions and Trajectories}
What originally motivated this research was the appearance and symmetries of computer simulated geodesics on the surface. These symmetries are related to the reflective and rotational symmetries of its translation cover.
\begin{figure}[H]
\centering
\includesvg[width=4in]{oddoddtrajectory}
\caption{A closed trajectory on the Necker cube surface modeled using sage-flatsurf \cite{flatsurf} and various sage utilities. [cite]}
\end{figure}
We began finding patterns between the choice of initial direction and periodicity. What we aimed to find out was what geometric properties of the surface relate a geodesic's dynamical characteristics to its initial trajectory angle. The answer to the question of what directions determine periodicity and drift periodicity turned out remarkably simple. Along the way we learned that the symmetries of this surface and its various covers gave up not only a clear distinction between periodic and drift-periodic trajectories, but the total arclengths of these paths after a single period. For example the previous figure's geodesic has an initial trajectory angle $\tan^{-1}(51/31)\approx58.71^\circ$, and total total length of $6\sqrt{3562}\approx 358.09$. Appendix [include appendix] showcases various geodesics, their initial trajectory angles, and total lengths.

\subsection{Discussion of Results}
A surface composed of infinitely many cubes is endowed with a flat metric and local plane isometries. Via the parallel transport of a unit vector tangent to the surface along a geodesic, we view its image as a sequence of straight lines on the smooth faces of the surface. (Fig 2) A geodesic is no doubt a piecewise curve in space. Let $\bUs$ be the connected, topological space known as the Necker cube surface.
\\\newpage
\begin{wrapfigure}{l}{1.9in}
\includesvg[width=1.7in]{vectorplane}
\end{wrapfigure}
\noindent Let $u_0$ be a point on the surface that is contained in the interior of a face and consider a tangent unit vector $v\in T_{u_0}(\bUs)$, the tangent vector field over $u_0$. A flattening map acts as an atlas on the smooth sections of $\bUs$ (see section [find section]) is an affine map from the surface to $\RR^2$ with a derivative in $\SO(3,\RR)$ that   (See section [?] in Appendix [Include section or appendix on this map].)
\\\\
Let the angle $v_0$ makes relative to a choice of basis (orthogonal vectors parallel to $\bUs$'s faces) be the \emph{initial trajectory angle} of a \emph{unit-speed geodesic} based at $u_0$, $\Phi_t:\RR\rightarrow\bUs$. The initial trajectory angle is said to be $\emph{rational}$ if there is some $k\in\RR$ such that $kv_0$'s nonzero components are relatively prime integers. A rational initial trajectory angle in $\RR^2$ falls into one of two categories:
\begin{Def}
Let $v_0$ be a rational unit vector of the form $\frac{1}{k}(x,y)\in\RR^2$ with $x,y\in\mathbb{Z}$ and $k=\sqrt{x^2+y^2}\in\mathbb{R}$. We say $kv_0$ is an \textbf{odd-odd} vector if $x,y$ are relatively prime and both odd. We denote the \textbf{set of all odd-odd directions} $\mathcal{O}$. We say that $kv_0$ is an \textbf{even-odd} vector if $x,y$ are relatively prime and of opposite parity. We denote the \textbf{set of all even-odd directions} $\mathcal{E}$. Finally, $\cS=\cO\sqcup\cE$ is the \textbf{collection of all possible rational directions}.
\end{Def}
\noindent It may be more helpful to think of an element in $\cS$ as the slope of $\Phi$, given $\bUs$'s flat structure. Recall that a geodesic on the surface is \emph{periodic} if there exists some $T>0$ such that $\Phi(t+T)=\Phi(t)$ for all $t\in\RR$, and \emph{drift-periodic} if there exists some $T'>0$ and non-trivial translation, $f$, such that $\Phi(t+T')=f(\Phi(t))$ for all $t\in\RR$. In any case we call this value the \emph{period} of $\Phi$ if it is the \emph{smallest} possible value that satisfies this. We additionally require that at no point does $\Phi$ encounter a cone singularity of the surface, where geodesic behavior is undefined. We call these well behaved geodesics \emph{non-singular}. We use $||~||:\RR^2\rightarrow\RR_{\geq 0}$ to denote the standard Euclidean norm in $\RR^2$.

\begin{thm*} \textbf{Periodic and Drift-Periodic Geodesics on the Necker cube surface.} 
\\Let $\Phi_t:\RR\rightarrow\bUs$ be a non-singular unit-speed geodesic on the Necker cube surface such that $\Phi(0)=u_0$ and $\Phi'(0)=v\in\RR^3$. Let $v_0\in\RR^2$ be a distance-preserving projection of $v$ onto the plane tangent to the surface at $u_0$. Let $k\in\RR$ such that $kv_0\in\cS$. Then the following is true:
\begin{enumerate}[label=(\roman*)]
\item $\Phi$ is periodic if $kv_0\in\cO$ with period $T=6||kv_0||$.
\item $\Phi$ is drift-periodic if $v_0\in\cE$ with period $T=2||kv_0||$.
\end{enumerate}
\end{thm*}

There is a very clear dynamical dichotomy at work when it comes to $\Phi$'s slope, not unlike the well-known Ehrenfest Wind-Tree model with $\frac{1}{2}\times\frac{1}{2}$ square obstacles distributed on a $\ZZ^2$ lattice. [cite]

@@

\section{Periodic Construction of the Surface}
This section will detail how the Necker Cube surface is constructed as a covering space of a thrice-punctured torus. We begin by making the following identifications on 3 unit squares:
\begin{figure}[H]
\centering
\includesvg[width=1.83in]{neckercube}\includesvg[width=2in.]{neckercubepath}
\caption{$\bGs$ with and without paths $a,b,c,d$.}
\end{figure}

We will call this ``half-cube" $\bG$. It is a piecewise smooth surface homeomorphic to the torus. Every vertex is a cone singularity of the surface with cone angle $3\pi$ or $\frac{3\pi}{2}$. Let $\Sigma_{\bG}\subset\bG$ be the set of these singularities. We use $\bG\backslash\Sigma_{\bG} = \bGs$ to denote the surface punctured at these points. The paths labeled $a,b,c,d$ are independent of one another and generate $\bGs$'s fundamental group, $\pi_1(\bGs)\cong \<a,b,c,d \>$, the free group of four generators. (Figure ?) The unpunctured Necker cube surface $\bU$ is embedded in $\RR^3$ by gluing along $\bG$'s outer edges (marked with solid triangles):

\begin{figure}[H]
\centering
\includesvg[width=4in]{neckercubepathsurface}
\end{figure}
From this diagram we see that $\bU$ is a universal cover of $\bG$. In this paper we are studying geodesic behavior on $\bU$ and, in order to do so properly, it is necessary to determine how the parallel transport of a unit vector around a cone singularity acts on the unit tangent bundle of the surface. But to study a given path's rotational holonomy, we require that these loops (see $c,d$ in Fig 3) aren't contractible through the cube's sharp corners. To that effect, we puncture the surface at these points.  Denote the countably infinite set of singularities of $\bU$ by $\Sigma_{\bU}$. The \emph{Necker cube surface} is the punctured surface $\bUs=\bU\backslash\Sigma_{\bU}$. Observe from the previous diagram that $\pi_1(\bUs)$ is the kernel of the following group homomorphism on $\pi_1(\bGs)$:

\begin{Def}
Let $\varphi_1:\pi_1(\bGs)\rightarrow \ZZ^2$ be the homomorphism on a free group given by $c,d\mapsto(0,0)$, $a\mapsto(1,0)$, and $b\mapsto(0,1)$.
\end{Def}

The cover of $\bGs$ with fundamental group $\ker\varphi_1$ is the space $\bUs=\tilde{\mathbf{G}}^\circ/\ker\varphi_1$ where $\ker\varphi_1$ acts on $\bGs$'s universal cover $\tilde{\mathbf{G}}^\circ$ by deck transformations.

\begin{Def}
Let $p_1:\bUs\rightarrow\bGs$ be a projection that induces a monomorphism on the fundamental groups of these surfaces such that $p_{1*}\pi_1(\bUs)=\ker\varphi_1$.
\end{Def}

It follows that $(\bUs,p_1)$ is a covering space of $\bGs$. We say $\Delta_{p_1}=\pi_1(\bGs)/p_{1*}\pi_1(\bUs)\cong\ZZ^2$ is the \emph{deck group} of the covering isomorphic to the group of deck transformations of $p_1$ $-~\text{Deck}_{p1}(\bUs)$. Using the fact that this cover is regular, we say that $p_1(\bUs)=\bUs/\ZZ^2$ and $\bGs$ is a quotient of the Necker cube surface. Apply the following cutting and gluing operations to $\bGs$: \\
\begin{figure}[H]
\centering
\includesvg[width=3in]{LTorus}
\end{figure}
\noindent This L-shaped torus is then taken apart and recovered as $2\times 2$ torus missing a unit square with the following edges identified:\\
\begin{figure}[H]
\centering
\includesvg[width=2in]{BranchTorus}
\caption{$\bGs$ and its paths after reconstruction.}
\end{figure}

Consider a $2\times2$ sheet missing an open unit square and that square's vertices (punctures). Center this structure around $0\in\CC$ and recover $\bGs$ with the identifications made in Fig ??. Let $m,n\in\ZZ$. Similar identifications can be made on the following subset of $\CC$ to recover $\bUs$:

@@

\begin{figure}[H]
\centering
\includegraphics[width=2.8in]{P.png}
\end{figure}
\compav{shade the surface}
\noindent Let $m,n\in\ZZ$. The coordinate $c_{m,n}=(2m+i2n)\in\CC$ is the center of a removed square. The set of all centers form a lattice in the plane that uniquely identifies every section in the fiber over $\bGs$ under $p_1$ with an element of $2\ZZ[i]$. Now let $\zeta_{m,n}(t)=(2m+t)+i(2n-\frac{1}{2})$ and 
$\xi_{m,n}(t)=(2m+t)+i(2n+\frac{1}{2})$ be paths on $\mathbf{P}$ that run along the top and bottom edges of the squares, parameterized by $t\in (-\frac{1}{2},\frac{1}{2})$ and $(m,n)\in\ZZ^2$. These paths are oriented so that edge identifications can be made by translations and rotations relative to $(m,n)$. (see figure above) $\bUs$ is recovered as the topological quotient $\mathbf{P}/\sim$, where $\sim$ is the following minimal relation on $\mathbf{P}$:

\begin{equation}
\begin{split}
\zeta_{m,n}(t)\sim i[\zeta_{m,n}(t)-c_{m,n}]+c_{m,n}-1\\
\xi_{m,n}(t)\sim i[\xi_{m,n}(t)-c_{m,n}]+c_{m,n}+1
\label{eq:rel2}
\end{split}
\end{equation}
The translations, $\pm 2,\pm 2i$, on $\mathbf{P}$ are isometries that preserve these identifications and induce isomorphisms of $\bUs$. We denote the isomorphism induced by $+2$ as $f_h$, for horizontal, and the isomorphism induced by $+2i$ as $f_v$, for vertical:

@@

From this image it's clear that the group generated by these horizontal and vertical isomorphisms under function composition, $\<f_h,f_v~\vert~[f_h,f_v] \>$, form $\text{Deck}_{p_1}(\bUs)$. Let $x_0\in\bGs$ and fix a point in the fiber, $u_0\in p_{1}^{-1}(x_0)\subset\bUs$. We consider the groups $\pi_1(\bUs,u_0)$ and $\pi_1(\bGs,x_0)$. There is a well-defined transitive action of the group $\pi_1(\bGs,x_0)$ on the set $p_{1}^{-1}(x_0)$ by lifting a loop on $\bGs$ to a curve on $\bUs$ based at a point in the fiber, and sending it to the terminal point of the lift. [cite that algebraic top book 5.7, beginning pg.133]. A curve is a continuous map from $[0,1]$ to the surface, and a loop is a closed curve where the terminal and initial point coincide.

\begin{thm}
Let $a,b\in\pi_1(\bGs,x_0)$ be the homotopy class generators as in Figure [??]. Let $\tilde{a},\tilde{b}$ be $u_0$-based lifts of a loop in $a,b$ onto $\bUs$. Then $\tilde{a}(1)=f_h(u_0)$ and $\tilde{b}(1)=f_v(u_0)$.
\end{thm}

\begin{cor}
Let $\alpha\in\pi_1(\bGs,x_0)$. Denote the unique $u_0$-based lift of a loop in $\alpha$ to $\bUs$ as $\tilde{\alpha}$. If $\varphi_1(\alpha)=(m,n)$, then 
\begin{align}
\tilde{\alpha}(1)=(f_h^m\circ f_v^n)(u_0).
\end{align}
\end{cor}
This just follows from the previous theorem as $c,d$ are mapped trivially, and under $\varphi_1$ a word in $\pi_1(\bUs)$ is taken to commutative automorphisms of $\bUs$. For shorthand, we will use $f_\alpha$ to denote the \emph{induced deck transformation.}

\begin{lem}
$f_\alpha$ is a local diffeomorphism of the surface $\bUs$. 
\begin{proof}
Consider a 
\end{proof}
\end{lem}

\subsection{Embedding the Necker cube surface}
In this section we define a map from $\bUs$'s $\RR^3$ embedded form to $\mathbf{P}$. The isomorphisms induced from translations on $\mathbf{P}$ are local diffeomorphisms on the smooth sections of $\bUs$, $S(\bUs)$, that relate the fundamental group of $\bGs$ to the tangent spaces of $p_1^{-1}(x_0)$. $S(\bUs)$ is the set of faces of cubes without boundary that are isometric to the plane. Let $m,n,l\in\ZZ$. A topologically equivalent notion of the (unpunctured) Necker cube surface is as the union of sets:  \\
\begin{align*}
\mathbf{A}=\bigcup\big{\{}\mathbf{A}_{m,n,l}: m+n+l=0\big{\}},
\\\mathbf{B}=\bigcup\big{\{}\mathbf{B}_{m,n,l}: m+n+l=0\big{\}},
\\\mathbf{C}=\bigcup\big{\{}\mathbf{C}_{m,n,l}: m+n+l=0\big{\}},
\end{align*}
which correspond to the following families of ``half-cubes" based integer points on a plane in $\RR^3$:

\vspace{0.2in}
\begin{tabular}{p{10cm}c}
\begin{align*}
\mathbf{A}_{m,n,l} = [m, m+1]\times[n,n+1]\times\{l\}, 
\\\mathbf{B}_{m,n,l} = \{m+1\}\times[n,n+1]\times[l-1,l],
\\\mathbf{C}_{m,n,l}= [m,m+1]\times\{n+1\}\times[l-1,l].
\end{align*}
&
\shiftleft{0.3in}{\raisebox{-1in}{\includegraphics[scale=1]{label.png}}}
\end{tabular}

\begin{figure}[H]
\centering
\includesvg[width=1in]{cubesxyzcut}
\caption{Embedding the Necker cube surface, i.e. $\bU=\mathbf{A}\cup\mathbf{B}\cup\mathbf{C}$.}
\end{figure}
